---
title: "Rats with Poisson GAMs"
output: github_document
---

```{r setup, include=FALSE}
library(dplyr)
library(ggplot2)
library(gratia)
load_mgcv()

source(here::here("gams", "gam_fxns", "fd_fxns.R"))
```


## The data
```{r show rat ts}
ts <- read.csv(here::here("gams", "rat_data.csv"))

ts_long <- ts %>%
  select(year, baileys, bannertail, merriami, pocketmouse) %>%
  tidyr::pivot_longer(-year, names_to = "species", values_to = "abundance") %>%
  filter(species != "total_abundance")

ggplot(ts_long, aes(year, abundance, color = species)) +
  geom_line() +
  theme_bw() +
  facet_wrap(vars(species))

```

## Baileys walkthrough

### Fit GAM and extract derivative
```{r fit gam and get derivatives}
pb_real_plot <- ggplot(ts, aes(year, baileys)) +
  geom_line()

pb_mod <- gam(baileys ~ s(year, k = 5), data = ts, method = "REML", family = "poisson")

pb_fit <- add_fitted(select(ts, year, baileys), pb_mod)

pb_fit_plot <- ggplot(pb_fit, aes(year, (baileys))) +
  geom_point() +
  geom_line(aes(year, .value))

pb_derivs <- get_many_fd(pb_mod, eps = .1)

pb_deriv_plot <- ggplot(pb_derivs, aes(year, derivative, group = seed)) +
  geom_line(alpha = .01) +
  geom_hline(yintercept = 0) +
  geom_line(aes(year, mean), color = "red") +
  geom_line(aes(year, upper), color = "red") +
  geom_line(aes(year, lower), color = "red")

gridExtra::grid.arrange(grobs = list(pb_real_plot, pb_fit_plot, pb_deriv_plot), ncol = 1)
```


### Summarizing derivatives

#### Net and absolute change
```{r net and absolute change}

pb_derivs <- pb_derivs %>%
  mutate(abs_derivative = abs(derivative)) %>%
  mutate(increment = derivative * eps,
         abs_increment = abs_derivative * eps)

pb_deriv_net <- sum(pb_derivs$increment)

pb_deriv_abs <- sum(pb_derivs$abs_increment)

pb_derivs_summary <- pb_derivs %>%
  group_by(seed) %>%
  summarize(net_change = sum(increment),
            abs_change = sum(abs_increment)) %>%
  mutate(abs_v_net_change = log(abs_change / net_change))

net_plot <- ggplot(pb_derivs_summary, aes(net_change)) +
  geom_density() +
  geom_vline(xintercept = 0)


abs_plot <- ggplot(pb_derivs_summary, aes(abs_change)) +
  geom_density() +
  geom_vline(xintercept = 0)


net_v_abs_plot <- ggplot(pb_derivs_summary, aes(abs_v_net_change)) +
  geom_density() +
  geom_vline(xintercept = 0)

gridExtra::grid.arrange(grobs = list(net_plot, abs_plot, net_v_abs_plot), nrow = 1)


```

From left to right, the net change over the TS, the absolute change (increasing + decreasing), and the log ratio of the absolute change to the net change.

* If the **net** change overlaps 0 to some confidence interval, an argument that we don't confidently have an overall increase or decrease over the full timeseries.
* If the **absolute** change **matches** the net change, whatever change occurred occurred in a consistent direction. The more the absolute change exceeds the net change, the more peaks and valleys occurred that may or may not be reflected in the net change.


#### Duration of change

```{r duration of change}

pb_derivs <- pb_derivs %>%
  group_by_all() %>%
  mutate(ci_sign = ifelse(all(upper < 0, lower < 0), "negative", 
                          ifelse(all(upper > 0, lower > 0), "positive", "zero"))) %>%
  ungroup()

pb_derivs_sign <- select(pb_derivs, year, mean, eps, ci_sign) %>%
  distinct()

sign_plot <- ggplot(pb_derivs_sign, aes(year, mean, color = ci_sign)) +
  geom_line(size = 5) +
  theme(legend.position = "none")

sign_barplot <- ggplot(pb_derivs_sign, aes(ci_sign, fill = ci_sign)) +
  geom_bar() 

gridExtra::grid.arrange(grobs = list(sign_plot, sign_barplot), nrow = 1)

pb_derivs_sign %>%
  group_by(ci_sign) %>%
  summarize(proportion_of_time = dplyr::n() / nrow(pb_derivs_sign),
            total_time = sum(eps))
```

<!-- ```{r nind over ts} -->
<!-- pb_derivs_sign <- pb_derivs_sign %>% -->
<!--   mutate(nind_over_ts = mean * eps * nrow(pb_derivs_sign)) %>% -->
<!--   mutate(greater_than_10 = abs(nind_over_ts) > 10) -->

<!-- ggplot(pb_derivs_sign, aes(year, mean, color = ci_sign)) +  -->
<!--   geom_line(size = 5, alpha = .2) +  -->
<!--   geom_line(data = filter(pb_derivs_sign, greater_than_10), aes(year, mean, color = ci_sign), size = 5, alpha = .8) -->


<!-- ``` -->

<!-- #### Gradual vs rapid change -->

<!-- I'm not sure how to do this. -->

<!-- ```{r variability of change} -->

<!-- ggplot(pb_derivs_sign, aes(abs(mean))) + -->
<!--   geom_density() + -->
<!--   geom_vline(xintercept = quantile(abs(pb_derivs_sign$mean), probs = c(.025, .1, .5, .9, .975, 1))) -->

<!-- ``` -->
