---
title: "Interpreting breakpoint fits"
output: github_document
params:
 this_dataset: ["all"]
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F)
library(dplyr)
library(ggplot2)
library(strucchange)
```

## An idea

Look specifically at the predictions from the best-fitting breakpoint model. (This model could be: no breakpoints + no slope, no breakpoints + slope, n breakpoints + no slope, n breakpoints + slopes).

There's some qualitative behaviors we can extract from the predictions (that come logically but less elegantly from the parameters, etc).

Two axes:

1. **Monotonic or squiggly:** Monotonic can encompass all linear (no-break) models, _and_ any models with breakpoints that do not result in changes in direction. 
2. **Net change or net 0:** The ratio of the end:beginning _prediction_.

The 2x2:

1. **Monotonic** and **net change**: This would be some kind of overall trend, either steady (probably would show as a 1-segment linear model), possibly accelerating or decelerating across the timeseries (would show as multiple segments with slopes), or even as a series of abrupt changes (would show as multiple segments with *no* slopes).
2. **Monotonic** and **no net change**: This can basically only be accomplished as a one-segment linear model with a slope very close to 0. (Theoretically you could have many verrrrry gently sloping segments, but I doubt we have the statistical power such that something like that would emerge as the best fit). I think this is either very stable or *so* variable that not even many breakpoints can adequately capture the variability. 
3. **Turnpoints** and **net change**: The TS changes direction at least once, and ends up somewhere other than where it started. This *must* involve breakpoints. At the moment I think I have less confidence in such an outcome as evidence of systematic directional change - it seems potentially sensitive to, if we stopped surveying 5 years earlier, would we have a totally different trend? 
4. **Turnpoints** and **no net change**: Qualitatively different from, monotonic and no net change....Again, I am not sure how confident I am in this as a signal of any kind of regulation/buffering. But maybe. 


### Illustration via a few real datasets

```{r loading some real datasets, fig.dim = c(6, 24)}

datasets <- data.frame(
  dataset_name = c("rockies",
                   "hartland",
                   "alberta",
                   "cochise_birds",
                   "salamonie",
                   "tilden",
                   #"gainesville",
                   #"gainesville_nooutlier",
                   "portal_rats",
                   "mccoy",
                   "cranbury"),
  rtrg_code = c("rtrg_304_17",
                "rtrg_102_18",
                "rtrg_105_4",
                "rtrg_133_6",
                "rtrg_19_35",
                "rtrg_172_14",
                #"rtrg_113_25",
                #"rtrg_113_25",

                NA,
                "rtrg_63_25",
                "rtrg_26_59")
)

if(params$this_dataset != "all") {
  datasets <- datasets %>%
    filter(dataset_name == params$this_dataset)

}
all_datasets <- list()


for(i in 1:nrow(datasets)) {

  if(datasets$dataset_name[i] != "portal_rats") {

    ibd <- readRDS(paste0("C:\\Users\\diaz.renata\\Documents\\GitHub\\BBSsize\\analysis\\isd_data\\ibd_isd_bbs_", datasets$rtrg_code[i], ".Rds"))

    sv <- ibd %>%
      group_by(year) %>%
      summarize(richness = length(unique(id)),
                abundance = dplyr::n(),
                biomass = sum(ind_size),
                energy = sum(ind_b)) %>%
      ungroup() %>%
      mutate(mean_energy = energy / abundance,
             mean_mass = biomass/abundance,
             site_name = datasets$dataset_name[i])

    if(datasets$dataset_name[i] == "gainesville_nooutlier") {
      sv <- filter(sv, abundance < 3000)
    }
  } else {

    individual_rats <- portalr::summarise_individual_rodents(clean = TRUE, type = "Granivores", time = "date", length = "Longterm")

    ibd <- individual_rats %>%
      filter(year %in% c(1978:2002), !is.na(wgt), treatment == "control") %>%
      mutate(six_mo = ifelse(month > 6, .5, 0)) %>%
      mutate(year_six_mo = (year + six_mo)) %>%
      mutate(bmr = 5.69 * (wgt ^ .75)) %>%
      select(year_six_mo, species, wgt, bmr) %>%
      rename(year= year_six_mo,
             id = species,
             ind_size = wgt,
             ind_b = bmr) %>%
      mutate(id = as.character(id))


    sv <- ibd %>%
      group_by(year) %>%
      summarize(richness = length(unique(id)),
                abundance = dplyr::n(),
                biomass = sum(ind_size),
                energy = sum(ind_b)) %>%
      ungroup() %>%
      mutate(mean_energy = energy / abundance,
             mean_mass = biomass/abundance,
             site_name = datasets$dataset_name[i]) %>%
      mutate(time = row_number())

  }

  all_datasets[[i]] <- sv


}

all_datasets <- bind_rows(all_datasets)
gridExtra::grid.arrange(grobs = list(
  ggplot(all_datasets, aes(year, abundance, color = site_name)) +
    geom_line() +
    theme_bw() +
    facet_wrap(vars(site_name), scales = "free", ncol = 1) +
    ggtitle("Abundance"
    ) +
    theme(legend.position = "top"),
  ggplot(all_datasets, aes(year, energy, color = site_name)) +
    geom_line() +
    theme_bw() +
    facet_wrap(vars(site_name), scales = "free", ncol = 1) +
    ggtitle("Energy") +
    theme(legend.position = "top")),
  ncol = 2
)

```

```{r breakpoints fxns}
source(here::here("stories", "datasets", "fxns.R"))

```

```{r multiple datasets}

abund_mods <- list()

for(i in 1:nrow(datasets)) {
  thisdat <- subset_all_datasets(site = datasets$dataset_name[i], curr = "abundance", all_datasets = all_datasets)

  abund_mods[[i]] <- summarize_breakpoints(thisdat, fit_breakpoints(thisdat))
}

abund_mods <- dplyr::bind_rows(abund_mods)

ggplot(abund_mods, aes(time, response, color = monotonic)) +
  geom_line() +
  geom_point(aes(time, fitted)) +
  facet_wrap(vars(site_name), scales = "free")



energy_mods <- list()

for(i in 1:nrow(datasets)) {
  thisdat <- subset_all_datasets(site = datasets$dataset_name[i], curr = "energy", all_datasets = all_datasets)

  energy_mods[[i]] <- summarize_breakpoints(thisdat, fit_breakpoints(thisdat))
}

energy_mods <- dplyr::bind_rows(energy_mods)

ggplot(energy_mods, aes(time, response, color = monotonic)) +
  geom_line() +
  geom_point(aes(time, fitted)) +
  facet_wrap(vars(site_name), scales = "free")

all_mods <- bind_rows(energy_mods, abund_mods) %>%
  mutate(site_curr = paste0(as.character(site_name), currency))
```

### Monotonic and net change

```{r monotonic net}

m_nc <- filter(all_mods, monotonic, net_change != 1)

ggplot(m_nc, aes(time, response)) +
  geom_line() +
  geom_point(aes(time, fitted)) +
  theme_bw() +
  facet_wrap(vars(site_curr), scales = "free")


```

### Monotonic and no net change

```{r monotonic no net}

m_nzero <- filter(all_mods, monotonic, net_change == 1) %>%
  group_by(site_curr) %>%
  mutate(cv = sd(response) / mean(response)) %>%
  ungroup()

ggplot(m_nzero, aes(time, response)) +
  geom_line() +
  geom_point(aes(time, fitted)) +
  theme_bw() +
  facet_wrap(vars(site_curr), scales = "free")


```

### Turnpoints and net change


```{r turns net, fig.dim = c(6,12)}

t_nc <- filter(all_mods, !monotonic, net_change != 1)

ggplot(t_nc, aes(time, response)) +
  geom_line() +
  geom_point(aes(time, fitted)) +
  theme_bw() +
  facet_wrap(vars(site_curr), scales = "free", ncol = 2)


```


<!-- ## Comparison to first/last 5 years -->


<!-- ```{r get first and last five years} -->


<!-- all_datasets_caps <- all_datasets %>% -->
<!--   group_by(site_name) %>% -->
<!--   mutate(time_step = row_number()) %>% -->
<!--   mutate(ntimesteps = max(time_step)) %>% -->
<!--   mutate(first_fifth = 5, -->
<!--          last_fifth = ntimesteps - 4) %>% -->
<!--   mutate(in_beginning = time_step <= first_fifth, -->
<!--          in_end = time_step >= last_fifth) %>% -->
<!--   mutate(in_cap = (in_beginning + in_end) > 0) %>% -->
<!--   mutate(which_cap = ifelse(in_beginning, 1, ifelse(in_end, 2, NA))) %>% -->
<!--   ungroup() -->

<!-- ggplot(all_datasets_caps, aes(year, abundance, color = in_cap)) + -->
<!--   geom_point() + -->
<!--   theme_bw() + -->
<!--   facet_wrap(vars(site_name), scales = "free") -->

<!-- ggplot(all_datasets_caps, aes(year, energy, color = in_cap)) + -->
<!--   geom_point() + -->
<!--   theme_bw() + -->
<!--   facet_wrap(vars(site_name), scales = "free") -->



<!-- all_caps <- lapply(unique(all_datasets$site_name), FUN = filter_caps, datasets_to_pass = all_datasets_caps) -->

<!-- all_caps_comparisons <- lapply(all_caps, FUN = compare_caps) -->


<!-- all_caps_energy <- lapply(unique(all_datasets$site_name), FUN = filter_caps, datasets_to_pass = all_datasets_caps, currency = "energy") -->

<!-- all_caps_energy_comparisons <- lapply(all_caps_energy, FUN = compare_caps) -->

<!-- all_caps_comparisons <- bind_rows(all_caps_comparisons) -->
<!-- all_caps_energy_comparisons <- bind_rows(all_caps_energy_comparisons) -->

<!-- all_caps_comparisons <- bind_rows(all_caps_comparisons, all_caps_energy_comparisons) -->

<!-- #all_datasets <- left_join(all_datasets, all_caps_comparisons) -->

<!-- ggplot(all_caps_comparisons, aes(which_cap, mean, color = pval < .05)) + -->
<!--   geom_point() + -->
<!--   geom_errorbar(aes(ymin = response_lower, ymax = response_upper)) + -->
<!--   facet_grid(rows = vars(currency), cols = vars(site_name), scales = "free") + -->
<!--   theme_bw() -->

<!-- ggplot(all_caps_comparisons, aes(which_cap, mean, color = wilcox_pval < .05)) + -->
<!--   geom_point() + -->
<!--   geom_errorbar(aes(ymin = response_lower, ymax = response_upper)) + -->
<!--   facet_grid(rows = vars(currency), cols = vars(site_name), scales = "free") + -->
<!--   theme_bw() -->

<!-- ``` -->

<!-- ### Net change via lm -->

<!-- ```{r select lm fit} -->

<!-- abund_slope_plots <- list() -->

<!-- for(i in 1:nrow(datasets)) { -->
<!--   thisdat <- subset_all_datasets(site = datasets$dataset_name[i], curr = "abundance", all_datasets = all_datasets) -->

<!--   abund_slope_plots[[i]] <- plot_lm_change(thisdat) -->
<!-- } -->

<!-- gridExtra::grid.arrange(grobs = abund_slope_plots) -->



<!-- energy_slope_plots <- list() -->

<!-- for(i in 1:nrow(datasets)) { -->
<!--   thisdat <- subset_all_datasets(site = datasets$dataset_name[i], curr = "energy", all_datasets = all_datasets) -->

<!--   energy_slope_plots[[i]] <- plot_lm_change(thisdat) -->
<!-- } -->

<!-- gridExtra::grid.arrange(grobs = energy_slope_plots) -->
<!-- ``` -->


<!-- ```{r select lm fit with p} -->

<!-- abund_slope_plots_p <- list() -->

<!-- for(i in 1:nrow(datasets)) { -->
<!--   thisdat <- subset_all_datasets(site = datasets$dataset_name[i], curr = "abundance", all_datasets = all_datasets) -->

<!--   abund_slope_plots_p[[i]] <- plot_lm_change(thisdat, use_p = T) -->
<!-- } -->

<!-- gridExtra::grid.arrange(grobs = abund_slope_plots_p) -->



<!-- energy_slope_plots_p <- list() -->

<!-- for(i in 1:nrow(datasets)) { -->
<!--   thisdat <- subset_all_datasets(site = datasets$dataset_name[i], curr = "energy", all_datasets = all_datasets) -->

<!--   energy_slope_plots_p[[i]] <- plot_lm_change(thisdat, use_p=T) -->
<!-- } -->

<!-- gridExtra::grid.arrange(grobs = energy_slope_plots_p) -->
<!-- ``` -->

